\LARGE \bf Human-Robot Interaction: Tackling the AI Challenges

### 2013-09-05 paper.tex
Many rewrite, fix, typos,...

Human-Robot interaction is an area filled with challenges for artifical
intelligence: dynamic, partially unknown environments that are not originally
designed for autonomous machines; a large variety of situations and objects to
deal with, with possibly complex semantics; physical interactions with humans
that requires fine, low-latency control, representation and management of
several mental models, good situation assessment skills...the list goes on.

@@

Human-robot interaction is a challenge for artificial intelligence. This field
indeed lays at the crossroad of several other domains of AI and requires to
tackle them in a holistic manner: Modeling humans and human cognition;
acquiring, representing, manipulating in a tractable way abstract knowledge;
reasoning on this knowledge to make decisions; and eventually instanciating
those decisions into physical actions in coordination with humans. Many AI
techniques are invited, from visual processing to symbolic reasoning, from task
planing to \emph{theory of mind} building, from reactive control to action
recognition and learning.

We do not tackle all these issues. This article attempts however to organize
them into a coherent challenge for artifical intelligence, and also explicit
some of the paths that have been investigated on our robots and that result in a
deliberative, knowledge-oriented, architecture for human-robot interaction.

@@

\begin {itemize}
\item a joint goal, which has been previously established and agreed
  upon (through a process which is out of the scope of this paper),
\item the current situation, acquired through perception or
  deduction from previous perceptions, including the state of the
  environment of the robot and of the human,
\item [\it a priori} common-sense knowledge,
\end {itemize}

@@

Previous section provided an overview of what we propose as a generic cognitive
architecture for robots involved in interactions with other social agents.

We now introduce each of the components and try to give insights on their main
features.

@@

\begin{figure}
        \centering
        \includegraphics[width=\columnwidth]{spark-oro}
    \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
    and the geometric situation assessment module (\emph{{\sc Spark}}, bottom
    part). {\sc Spark} computes symbolic relationships between objects and
    agents, and fed them to the knowledge base.}

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module called {\sc Spark} (\emph{SPAtial Reasoning \&
Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
generates symbolic knowledge from the geometry of the environment with respect
to relations between objects, robots and humans (Figure~\ref{fig|spark-oro}),
also taking into account the different perspective that each agent has on the
environment.

{\sc Spark} is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
environment that serves both as basis for the fusion of the perception
modalities and as bridge with the symbolic layer. This geometric model is
continuously updated at run-time by the robot based on its sensors (in our
experiments, objects are identified and localised through 2D barcodes, while
humans are tracked with Kinect-like devices, optionally assisted by motion
capture to accurately track the head motion, which is required to compute what
the human is looking at).

\fxnote{Add 2-3 lines to summarize the AI highlights of {\sc Spark}}

@@

Figure~\ref{fig::reach-ex} illustrates different situations for the
\textit{reachable} relation. The robot and its human partner are placed face to
face (Figure~\ref{fig::reach-ex}.1).  The robot first estimates if the small
grey box is reachable to itself. This is done by finding a collision free
posture to reach the object (Figure~\ref{fig::reach-ex}.2). Next the robot
switches to the human's perspective to estimate if the same object is reachable
to the human as well.  In the last scene, the human moves towards his left,
farther from the object (Figure~\ref{fig::reach-ex}.4). The situation is then
reevaluated. In this occasion though, the reasoner cannot find a satisfactory
posture for the human to reach the box because he is too far from the target.

\begin{figure*}[!t]
	\centering
    \subfigure[] { \includegraphics[width=0.2\textwidth]{reachex-1.jpg} }
    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-2.jpg} }
    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-3.jpg} }
    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-4.jpg} }
	\caption{An example illustrating the \textit{reachable} relation. The relation is computed from the perspectives of both the robot and the human. The computed posture at each step is illustrated with a global view of the scene (top), and from a closest view (bottom).}
\label{fig::reach-ex}
\end{figure*}

\begin{figure*}[ht!]
   \begin{center}
%
       \subfigure[Initial state]{%
%           \label{fig:wuweiPhoto}
           \includegraphics[width=0.5\textwidth]{etat2-P1010769_brightened-v2.jpg}
       }%
       \subfigure[3d model view of initial state]{%
%          \label{fig:sparkScreenshot}
          \includegraphics[width=0.43\textwidth]{etat2_photo.png}
       }\\ %  ------- End of the first row ----------------------%
%
   \end{center}

   \caption{%
     In this situation, there are three tapes on the table. Two tapes
     are only reachable by the robot: the LOTR\_TAPE (black in the 3d
     model) and GREY\_TAPE. The third tape WALLE\_TAPE (white in the
     3d model) and the trashbin PINK\_TRASHBIN are only reachable by
     the human HUMAN1. All tapes are on the table TABLE.  }%
   \label{fig:sparkSubfigures}

The set of facts computed in the situation depicted by
Figure~\ref{fig:sparkSubfigures} is the following:
\begin{footnotesize}
%\begin{small}
\begin{verbatim}
                ROBOT                          HUMAN1
PINK_TRASHBIN isReachable false    PINK_TRASHBIN isReachable true 
WALLE_TAPE isReachable false       WALLE_TAPE isVisible true 
LOTR_TAPE isReachable true         LOTR_TAPE isReachable false 
GREY_TAPE isReachable true         GREY_TAPE isReachable false
WALLE_TAPE isVisible true          WALLE_TAPE isReachable true 
LOTR_TAPE isVisible true           LOTR_TAPE isVisible true 
GREY_TAPE isVisible true           GREY_TAPE isVisible true 
WALLE_TAPE isOn TABLE              WALLE_TAPE isOn TABLE 
LOTR_TAPE isOn TABLE               LOTR_TAPE isOn TABLE 
GREY_TAPE isOn TABLE               GREY_TAPE isOn TABLE 
\end{verbatim}
%\end{small}
\end{footnotesize}

\paragraph{Hypotheses on objects states and positions}
It is sometimes difficult or even impossible to see and/or track an
object in certain states. This happens, for instance, when the object
has been put in a container, when it is in the robot gripper or in the
human hand, and more generally in any state in which it is hidden by
something else. Our robot has a model of the possible symbolic states
for an object (whether the object is on a furniture, in an agent hand,
in a container, etc.).  According to the robot perception of what has
happened since the object was last seen, the robot tries to maintain a
belief of the current possible symbolic states and their associated
probabilities for this object. Such information can be used to update
the beliefs using input from exploration, dialog, human visual
focus,\ldots \fxnote{Paragraph need write. Tell what the module actually does, not what it could do}

{\sc Spark} currently provides a simple implementation of such a functionality. The
only managed hypotheses are \emph{in container} and \emph{in agent hand}. We
can have only one hypothesis at the same time. Hypothesis validity is checked
geometrically in case of incoming perception values.

@@

{\sc Spark} enables \emph{perspective-taking}: spatial relations between entities can
be computed from different viewpoints, which let the robot build for each
agents it interacts with a different, \emph{perspective-aware} symbolic model
of the environment. These models are separately stored in the knowledge base.
\fxwarning{Rephrase: already mentioned before}

@@

\subsubsection{Symbolic facts and beliefs management}
\fxwarning{Do we need this section here?}

The facts produced by the geometric and temporal reasoning component
are stored in the central symbolic knowledge base.

ORO stores independent knowledge models (in our implementation, as
\emph{ontologies}) for each agent (the robot and the humans it
interacts with). The robot architecture components (like the executive
layer or the situation assessment component) can then store the
agents' beliefs in specific models.  Each of these models is
independent and logically consistent, enabling reasoning on different
perspectives of the world that would otherwise be considered as
globally inconsistent (for instance, an object can be visible for the
robot but not for the human. This object can have at the same time the
property \concept{isVisible \textbf{true}} and \concept{isVisible
  \textbf{false}} in two different models). This feature actually
allows us to consider the robot to be endowed with a simple
\emph{theory of mind}~\cite{Scassellati2002}: it can explicitly
model the belief state of its interactors.

@@

Natural language input (in experiments, we rely on an Android-based interface,
with Google speech recognition) is parsed into a grammatical structure, and
atoms of each sentence are resolved with the help of the ontology to ground
concepts like objects (\ie when a user says ``pick the can'', resolve to which
instance of \emph{can} the user is referring to) and actions. Sentences are
sorted into questions, desires and statements, and processed accordingly.\fxwarning{Severin: Develop this section}

The system supports quantification (``give me \{a | the | some | all | any |
...\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to th eknowledge pool),
interprets common temporal and place adverbs (like \emph{above} or
\emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm
tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
hasFeature tired}).

\fxwarning{Give more implementation details. Why did we go for a custom NLP? (Severin)}

\subsubsection{Limits of disambiguation at semantic level}

@@

Because all components rely on the same RDF formalism to format their outputs,
the different communication modalities (\emph{explicit} like verbal, deictic or
based on gaze, or \emph{implicit} like postures) are presented in a homogeneous
way. The dialogue grounding process makes use of them at two distinct levels.

@@

HATP originality resides in its ability to produce plans for the robot's
actions \emph{as well as for the other participants} (humans or robots): the
robot plans for itself, but also for its interactors, which effectively endows the robot
with new interaction capabilities: It can partially predict the behaviour, and
thus monitor the engagement, of its human partner; the planner generates
synchronisation points (arrow going from the robot stram to the human stream on
Figure~\ref{plan_hatp1}) between the agents; the controller can also verbalize
the plans to explain its partner how a task may be shared~\cite{warnier2012when}.

@@

\emph{Desires} are typically human orders (``Give me that book''). The nature
of the desired action (to pick, give, look, bring, show...), along with the
action parametrization (what is acted on? who should perform the action? etc.)
are extracted from the knowledge base, and either passed to a task planner
(presented in the next section) or executed if the procedure is directly
available.

\emph{Experiences}, on the other hand, comprise of emotions, states and
questions (when asking a question, we consider the human to be in an
\emph{interrogative state}). When the knowledge base recognizes that an agent
\emph{experiences} a particular emotion or state, the execution controller may
decide to handle it, typically by trying to answer the question or using the
emotional or physical state as a parameter for subsequent actions. As an example,
when the speaker says ``I feel tired'', we change the motion planner
parametrization to lower the effort the human needs to provide for the following
joint manipulation tasks. Note that this example has been implemented as a
proof-of-concept. We have not yet tried to define a theoretical framework that
would support action alteration based on the user's experienced states.

@@

The {\sc Oro} server proposes two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. In its simplest
form, a module can subscribe to an event by passing through an event pattern
(in its basic form, a partial statement like \stmt{? type Elephant}) and a
callback.  Each time a new instance of elephant appears in the knowledge base,
the callback is triggered.

@@

Our other controller, {\sc Shary}, relies on the symbolic task planner to
generate at runtime suitable tasks.

@@

\paragraph{Human aware motion, placement and manipulation planning}
Manipulation and navigation actions listed in
table~\ref{table|pyrobots_actions} rely on a collection of dedicated 3D motion
planner to generate movements that are aware of surrounding humans. Robot
placement and end-effectors trajectories are computed on-line to allow object
manipulation that take into account task specific constraints and human
postures, abilities and preferences. \cite{Sisbot2008, Mainprice2011,
Pandey2010} detail these techniques.

\paragraph{Action Execution and Monitoring Robot Controller}
context and on the shared plan elaborated by HATP for a given goal,
the robot controller decides to execute an action or to ask its human
partner to do it.  Actions feasibility by the human or the robot are
regularly reconsidered based on the reachability / visibility
computation mechanisms.

@@

For plan action monitoring, primitive actions recognition defined
above are used. A primitive action detection is interpreted as action
success if it is the expected one and failure otherwise. The robot
also reacts to the absence of activity.

\paragraph{Context Management Policies} \fxfatal{Rewrite this section to level up AI aspects (Mathieu)}
Building, updating and maintaining a correct state of the world at
geometric and symbolic level is crucial to the capacity of the robot
to carry on successfully a multi-step interaction with a human. Tight
integration between the robot controller and the geometric and
temporal reasoning functions in {\sc Spark} and symbolic facts and beliefs
management in ORO is central.

The robot controller has access to the symbolic facts in ORO that are
automatically updated whenever object and agent positions are changed.
Robot controller can also access geometric perceived or inferred
positions of objects and geometric positions and postures of the human
that will be used to orient its cameras.  Building and updating the
state of the world first relies on perceiving objects. Robot
controller can use:

\begin {itemize}
\item Exploration policies: robot will exhaustively scan the table to
  see all what can be seen.

\item Search policies: robot will search an object until it is
  detected if possible, scanning all the table and looking in human
  hand.

\end {itemize}

Robot reasons on possible positions for non perceived objects. These hypotheses
could be updated\fxwarning{are they?} using new input from dialogue, human action and focus of
attention. Currently, we manage at most one hypothesis per object. This
hypothesis is produced by robot controller through an inference on robot or
human action. In case of perception conflicts with low probability for the
current hypothesis, robot controller will break this hypothesis and delete
corresponding symbolic fact in the ontology.

Robot must reacts to change in the world not linked to robot action to drive
world update. Robot controller uses {\sc Spark} to monitor human hand motion and
primitive action recognition for \emph{pick} and \emph{throw}. As mentioned
above these primitive action recognition should be used with higher level
information on goals, intentions and plans and some exploration to achieve
complex human action and activity monitoring.  In the current implementation,
these primitive actions are over-optimistically interpreted as the
corresponding actions \emph{pick} object and \emph{throw} object.

@@

The second one involves the SHARY execution controller and the HATP symbolic
task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
robot try to cooperatively remove objects from a table. In this scenario,
behaviours and motions are fully planned.

@@

This first experiment is based on the following daily life situation: Tom and
Jerry are moving to London, so they are packing things in boxes. The scenario
takes places in the living-room, where Jido (our robot) is observing while they
move things here and there. To assess the reasoning abilities of the robot they
ask Jido for information (entered through keyboard). Ideally, the robot should
also perform actions when required (e.g. hand an object when asking ``give
me...''). However, since it is out of the scope of this work, we do not include
any motion from the robot's side.

Perception of objects is done through a tag-based system and humans are
detected through motion capture. The robot knowledge base is pre-loaded with
the \emph{ORO Commonsense Ontology}.  We next describe in detail two
situations where we can follow the internal robot's reasoning and the
interaction with the users.

@@

Tom enters the room while carrying a big box (Figure~\ref{fig|vpt}, page 1). He
approaches the table and asks Jido to handle him the video tape: ``Jido, can
you give me the video tape''. The \textsc{Dialogs} module queries the ontology to
identify the object the human is referring to: \stmt{?obj type VideoTape}.

@@

However, only one is visible for Tom (the one on the
table). Thus, although there is an ambiguity from the robot's perspective
(since it can see both video tapes), based on the perspective of its human
partner it infers that Tom is referring to the video tape on the table, and not
the one inside the box which is not visible from his view. Therefore,
non-visible objects are removed obtaining: \concept{?obj =[videoTape1]}.

Since only one object is available, the robot infers
that the human refers to it and would eventually execute the command, \ie give
it to the human. Alternatively, the robot could first verify with the human if
that was the object being referred to or not before proceeding to execute the
action. Table~\ref{table|ptbeliefs} lists the robot's beliefs about itself and
its human partner involved in this situation.

@@

In this situation, Jerry enters the
living room without knowing where Tom had placed the video tapes. So he first
asks Jido: ``What's in the box?''. Before the robot can answer the question it
has to figure out which box Jerry is talking about. Similar to the previous
situation, there are two available boxes:

@@

However both are visible and the cognitive ambiguity resolution cannot be
applied. The only option is to ask Jerry which box he is referring to: ``Which
box, the toolbox or the cardboard box?'' Jerry could now simply answer the
question. Instead, he decides to point at it while indicating: ``This box''
(Figure~\ref{fig|pointing}). The robot's perception identifies the {\tt
cardBoardBox} as being pointed at and looked at by the human and updates the
ontology with this new information using a rule available in the commonsense
ontology ({\tt \textbf{pointsAt}(?ag, ?obj) $\land$ \textbf{looksAt}(?ag, ?obj) $\to$
\textbf{focusesOn}(?ag, ?obj)}) The \textsc{Dialogs} module is then able to merge both
sources of information, verbal (``this'') and gestural to distinguish the box
Jerry refers to.

@@

At this point Jerry wants to know where the other tape is, and that is exactly
what he asks Jido: ``And where is the other tape?''. In this occasion, the
\textsc{Dialogs} module is able to interpret that Jerry is not referring to the
video which they were just talking about, but to the other one:

@@

This experiment involved a more complex decisional layer where the ORO
server was used in conjunction with the HATP symbolic task planner and the
SHARY execution controller (the complete software architecture we deployed for
this experiment is pictured on Figure~\ref{fig|archi},
page~\pageref{fig|archi}).

In this scenario (Figure~\ref{fig|cleantable-video}), a human and a robot
cooperate to remove objects from a table. The robot produces symbolic plans for
both itself and the human (Figure~\ref{plan-etat2})
that allow the robot to (verbally) share the task with the human (like ``I take
the green box and I put it in the trashbin, you take the black video tape and
you throw it in the trashbin''). Plans are created based on perceived
visibility and reachability of the objects, and the robot also monitors the
human activities to track the advancement of the whole plan (this last aspect
is presented in~\cite{warnier2012when}).

@@

Figure~\ref{fig|cleantable-timeline} presents a excerpt of the whole task and
shows how the different components produce and use symbolic knowledge (note
that the plan is actually computed {\it a priori}. It appears nevertheless on
this diagram for better readability).

@@

\begin{figure*}[thpb]
  \centering
    \includegraphics[width=1.0\textwidth]{manip_run.pdf} \\
    \begin{center}
        \caption {Sequence of events of a simplified version of the \emph{Cleaning the table} scenario.}
    \end{center}
  \label{fig|cleantable-timeline}
\end{figure*}

>>>>

Human-Robot interaction is an area filled with challenges for artificial
intelligence: dynamic, partially unknown environments that are not originally
designed for autonomous machines; a large variety of situations and objects to
deal with, with possibly complex semantics; physical interactions with humans
that requires fine, low-latency control, representation and management of
several mental models, good situation assessment skills...the list goes on.

@@

Human-robot interaction is a challenge for artificial intelligence. This field
indeed lays at the crossroad of several other domains of AI and requires to
tackle them in a holistic manner: Modeling humans and human cognition;
acquiring, representing, manipulating in a tractable way abstract knowledge;
reasoning on this knowledge to make decisions; and eventually instantiating
those decisions into physical actions in coordination with humans. Many AI
techniques are invited, from visual processing to symbolic reasoning, from task
planing to \emph{theory of mind} building, from reactive control to action
recognition and learning.

We do not tackle all these issues. This article attempts however to organize
them into a coherent challenge for artificial intelligence, and also explicit
some of the paths that have been investigated on our robots and that result in a
deliberative, knowledge-oriented, architecture for human-robot interaction.

@@

\begin {itemize}
\item a joint goal, which has been previously established and agreed
  upon (through a process which is out of the scope of this paper),
\item the current situation, acquired through perception or
  deduction from previous perceptions, including the state of the
  environment of the robot and of the human,
\item {\it a priori} common-sense knowledge,
\end {itemize}

@@

Previous section provided an overview of our cognitive architecture for robots
involved in interactions with other social agents, as well as the knowledge
model we rely on.

We now introduce each of the components and give insights on their main
features.

@@

\begin{figure}
        \centering
        \includegraphics[width=\columnwidth]{spark-oro.png}
    \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
    and the geometric situation assessment module (\emph{{\sc Spark}}, bottom
    part). {\sc Spark} computes symbolic relationships between objects and
    agents, and feeds them to the knowledge base.}

@@

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module called {\sc Spark} (\emph{SPAtial Reasoning \&
Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
generates symbolic knowledge from the geometry of the environment with respect
to relations between objects, robots and humans (Figures~\ref{fig|spark-oro} and~\ref{fig:sparkSubfigures}),
also taking into account the different perspective that each agent has on the
environment.

{\sc Spark} is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
environment that serves both as basis for the fusion of the perception
modalities and as bridge with the symbolic layer. This geometric model is
continuously updated at run-time by the robot based on its sensors (in our
experiments, objects are identified and localised through 2D fiducial markers, while
humans are tracked with Kinect-like devices, optionally assisted by motion
capture to accurately track the head motion, which is required to compute what
the human is looking at).

From an artificial intelligence perspective, {\sc Spark} takes care of merging
sensing modalities, it computes and grounds perceptions into symbolic
knowledge (situation assessment), it is responsible temporal interpretation of
situations and performs spatial perspective-taking.

\begin{figure*}[ht!]
   \begin{center}
%
       \subfigure[Initial state]{%
%           \label{fig:wuweiPhoto}
           \includegraphics[width=0.5\textwidth]{etat2-P1010769_brightened-v2.jpg}
       }%
       \subfigure[Corresponding 3D model view]{%
%          \label{fig:sparkScreenshot}
          \includegraphics[width=0.43\textwidth]{etat2_photo.png}
       }\\ %  ------- End of the first row ----------------------%
%
   \end{center}

   \caption{% In this situation, there are three tapes on the table. Two tapes
       are only reachable by the robot: the \concept{LOTR\_TAPE} (black in the 3D
       model) and \concept{GREY\_TAPE}. The third tape \concept{WALLE\_TAPE} (white
       in the 3D model) and the trashbin \concept{PINK\_TRASHBIN} are only
       reachable by the human \concept{HUMAN1}. All tapes are on the table
   \concept{TABLE}. The corresponding symbolic facts are listed in Table~\ref{table|beliefsfig7}.  }%

 \label{fig:sparkSubfigures}

\begin{table}
\begin{center}
\begin{tabular}{l}
\hline
Robot's beliefs about itself (\emph{robot's model}):\\
\hline
  \hspace{0.7cm}\stmt{PINK\_TRASHBIN isReachable false}\\
  \hspace{0.7cm}\stmt{WALLE\_TAPE isReachable {\bf false}}\\
  \hspace{0.7cm}\stmt{LOTR\_TAPE isReachable {\bf true}}\\
  \hspace{0.7cm}\stmt{GREY\_TAPE isReachable {\bf true}}\\
  \hspace{0.7cm}\stmt{WALLE\_TAPE isVisible true}\\
  \hspace{0.7cm}\stmt{LOTR\_TAPE isVisible true}\\
  \hspace{0.7cm}\stmt{GREY\_TAPE isVisible true}\\
  \hspace{0.7cm}\stmt{WALLE\_TAPE isOn TABLE}\\
  \hspace{0.7cm}\stmt{LOTR\_TAPE isOn TABLE}\\
  \hspace{0.7cm}\stmt{GREY\_TAPE isOn TABLE}\\
\hline
\hline
Robot's beliefs about the human (\emph{human's model}):\\
\hline
  \hspace{0.7cm}\stmt{PINK\_TRASHBIN isReachable false}\\
  \hspace{0.7cm}\stmt{WALLE\_TAPE isReachable {\bf true}}\\
      \hspace{0.7cm}\stmt{LOTR\_TAPE isReachable {\bf false}}\\
      \hspace{0.7cm}\stmt{GREY\_TAPE isReachable {\bf false}}\\
  \hspace{0.7cm}\stmt{WALLE\_TAPE isVisible true}\\
  \hspace{0.7cm}\stmt{LOTR\_TAPE isVisible true}\\
  \hspace{0.7cm}\stmt{GREY\_TAPE isVisible true}\\
  \hspace{0.7cm}\stmt{WALLE\_TAPE isOn TABLE}\\
  \hspace{0.7cm}\stmt{LOTR\_TAPE isOn TABLE}\\
  \hspace{0.7cm}\stmt{GREY\_TAPE isOn TABLE}\\ 
 \hline
\end{tabular}
\end{center}
\caption{Symbolic facts computed from the situation depicted in
Figure~\ref{fig:sparkSubfigures}. Note how reachability differs for the two
agents.}

\label{table|beliefsfig7}
\end{table}

@@

Figure~\ref{fig::reach-ex} illustrates different situations for the
\textit{reachable} relation.

\begin{figure*}[!t]
	\centering
    \subfigure[] { \includegraphics[width=0.2\textwidth]{reachex-1.png} }
    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-2.jpg} }
    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-3.jpg} }
    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-4.jpg} }
    \caption{An example illustrating the \textit{reachable} relation. The
        relation is computed from the perspectives of both the robot and the
        human. The computed posture at each step is illustrated with a global
        view of the scene (top), and from a closest view (bottom). The robot
        and its human partner are placed face to face (a).  The robot first
        estimates if the small grey box is reachable to itself. This is done by
        finding a collision free posture to reach the object (b). Next the
        robot switches to the human's perspective to estimate if the same
        object is reachable to the human as well (c).  In the last scene, the
        human moves towards his left, farther from the object (d). The
        situation is then reevaluated. In this occasion though, the reasoner
        cannot find a satisfactory posture for the human to reach the box
        because he is too far from the target.  }

\label{fig::reach-ex}
\end{figure*}

\paragraph{Hypotheses on objects states and positions}
It is sometimes difficult or even impossible to see and/or track an
object in certain states. This happens, for instance, when the object
is hidden in a container, when the robot holds it in its gripper, and more
generally in any state in which it is hidden by something else. {\sc Spark} models the possible symbolic states of objects (whether the object is on a furniture, in an agent hand,
in a container, etc.): According to the robot perception of what has
happened since the object was last seen, the robot builds a
belief of the current possible state with associated probability for the.

@@

{\sc Spark} also supports \emph{perspective-taking}: spatial relations between entities can
be computed from different viewpoints, which let the robot build for each
agents it interacts with a different, \emph{perspective-aware} symbolic model
of the environment. These models are separately stored in the knowledge base.

@@

@@

Natural language input (in experiments, we rely on an Android-based interface,
with Google speech recognition) is parsed into a grammatical structure, and
atoms of each sentence are resolved with the help of the ontology to ground
concepts like objects (\ie when a user says ``pick the can'', it resolves to which
instance of \emph{can} the user is referring to) and actions. Sentences are
sorted into questions, desires and statements, and processed accordingly.
Figure~\ref{dialogs|ex} gives a example of the processing of a simple order.

\begin{figure}
    \centering
	\begin{tabular}{l|l}
	\emph{Initial human knowledge} &
	\emph{Human input}\\

	\hline

    	\stmt{book\_1 type Book} &
	``Give me the book.'' \\

    	\stmt{book\_1 hasColor blue} & \\
	\vspace{0.5em}\\
	\hline

	\emph{Generated partial statements} &
	\emph{Newly created statements}\\
	\hline
    	\stmt{?obj type Book} & 
	\stmt{human desires sit\_1} \\

	\hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1}
    	& \stmt{sit\_1 type Give} \\
    	& \stmt{sit\_1 performedBy myself} \\
    	& \stmt{sit\_1 actsOnObject book\_01} \\
    	& \stmt{sit\_1 receivedBy human} \\
	\end{tabular}

    \caption{Processing an order: this simple example (no ambiguity), taken
    from~\cite{Lemaignan2011a}, shows the thematic roles computed by {\sc
Dialogs} for the order ``Give me the book''.}

\label{dialogs|ex}
\end{figure}

The system supports quantification (``give me \{a | the | some | all | any |
...\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to the knowledge base),
interprets common temporal and place adverbs (like \emph{above} or
\emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm
tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
hasFeature tired}).

@@

Because all components rely on the same RDF formalism to format their outputs,
the different communication modalities (\emph{explicit} like verbal, deictic or
based on gaze, or \emph{implicit} like postures) are presented in a homogeneous
way, as statements in the knowledge base. The dialogue grounding process makes
use of them at two distinct levels to provide seamless multi-modal interaction.

@@

HATP originality resides in its ability to produce plans for the robot's
actions \emph{as well as for the other participants} (humans or robots): the
robot plans for itself, but also for its interactors, which effectively endows the robot
with new interaction capabilities: It can partially predict the behaviour, and
thus monitor the engagement, of its human partner; the planner generates
synchronisation points (arrow going from the robot stream to the human stream on
Figure~\ref{plan_hatp1}) between the agents; the controller can also verbalize
the plans to explain its partner how a task may be shared~\cite{warnier2012when}.

@@

\emph{Desires} are typically human orders (``Give me that book''). The nature
of the desired action (to pick, give, look, bring, show...), along with the
action parametrization (what is acted on? who should perform the action? etc.)
are extracted from the knowledge base, and either passed to a task planner
(presented in the previous section) or executed if the procedure is directly
available.

\emph{Experiences}, on the other hand, comprise of emotions, states and
questions (when asking a question, we consider the human to be in an
\emph{interrogative state}). When the knowledge base states that an agent
\emph{experiences} a particular emotion or state, the execution controller may
decide to handle it, typically by trying to answer the question or using the
emotional or physical state as a parameter for subsequent actions. As an
example, when the speaker says ``I feel tired'', we change the motion planner
parametrization to lower the effort the human needs to provide for the
following joint manipulation tasks. Note that this example has been implemented
as a proof-of-concept. A broader framework that would support action alteration
based on the user's experienced states is yet to be devised.

@@

The {\sc Oro} server proposes two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. A module can
subscribe to an event by passing through an event pattern (in its simplest
form, a partial statement like \stmt{? type Elephant}) and a callback.  Each
time a new instance of elephant appears in the knowledge base, the callback is
triggered.

@@

Our other controller, {\sc Shary}, relies on the symbolic task planner to
generate at runtime suitable sets of atomic actions.

@@

\paragraph{Human aware motion, placement and manipulation planning}
Manipulation and navigation actions listed in
table~\ref{table|pyrobots_actions} rely on a collection of dedicated 3D motion
planner to generate movements that are aware of surrounding humans. Robot
placement and end-effectors trajectories are computed on-line to allow object
manipulation that take into account task specific constraints and human
postures, abilities and preferences. \cite{Sisbot2008, Mainprice2011,
Pandey2011} detail these techniques.

\paragraph{Action Execution and Monitoring Robot Controller}
Based on context and on the shared plan elaborated by HATP for a given goal,
the robot controller decides to execute an action or to ask its human
partner to do it.  Actions feasibility by the human or the robot are
regularly reconsidered based on the reachability / visibility
computation mechanisms.

@@

For plan action monitoring, primitive actions recognition is used. A primitive
action detection is interpreted as action success if it is the expected one and
failure otherwise. The robot also reacts to the absence of activity.

@@

The second one involves the SHARY execution controller and the HATP symbolic
task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
robot try to cooperatively remove objects from a table. Behaviours and motions
are fully planned.

@@

This first experiment is based on the following daily life situation: Tom and
Jerry are moving to London, so they are packing things in boxes. The scenario
takes places in the living-room, where Jido (our robot) is observing while they
move things here and there. To assess the reasoning abilities of the robot they
ask Jido for information. Ideally, the robot should
also perform actions when required (\eg hand an object when asking ``give
me...'').

Perception of objects is done through a tag-based system and humans are
detected through motion capture. The robot knowledge base is pre-loaded with
the \emph{ORO Commonsense Ontology}.  We next describe two
situations where we can follow the internal robot's reasoning and the
interaction with the users.

@@

Tom enters the room while carrying a big box (Figure~\ref{fig|vpt}). He
approaches the table and asks Jido to hand him the video tape: ``Jido, can
you give me the video tape''. The \textsc{Dialogs} module queries the ontology to
identify the object the human is referring to: \stmt{?obj type VideoTape}.

@@

However, only one is visible for Tom (the one on the
table). Thus, although there is an ambiguity from the robot's perspective
(since it can see both video tapes), based on the perspective of its human
partner it infers that Tom is referring to the video tape on the table, and not
the one inside the box which is not visible for him.

Since only one object remains, the robot infers
that the human refers to it and would eventually execute the command, \ie give
it to the human. Alternatively, the robot could first verify with the human if
that was the object being referred to or not before proceeding to execute the
action. Table~\ref{table|ptbeliefs} lists the robot's beliefs about itself and
its human partner involved in this situation.

@@

In this situation, Jerry enters the living room without knowing where Tom had
placed the video tapes (Figure~\ref{fig|pointing}). So he first asks Jido:
``What's in the box?''. Before the robot can answer the question it has to
figure out which box Jerry is talking about. Similar to the previous situation,
there are two available boxes:

@@

However both are visible and the cognitive ambiguity resolution cannot be
applied. The only option is to ask Jerry which box he is referring to: ``Which
box, the toolbox or the cardboard box?'' Jerry could answer the
question. Instead, he decides to point at it while indicating: ``This box''
(Figure~\ref{fig|pointing}). The robot's perception identifies the {\tt
cardBoardBox} as being pointed at and looked at by the human and updates the
ontology with this new information using a rule available in the commonsense
ontology ({\tt \textbf{pointsAt}(?ag, ?obj) $\land$ \textbf{looksAt}(?ag, ?obj) $\to$
\textbf{focusesOn}(?ag, ?obj)}) The \textsc{Dialogs} module is then able to merge both
sources of information, verbal (``this'') and deictic to distinguish the box
Jerry refers to.

@@

At this point Jerry wants to know where the other tape is, and that is exactly
what he asks Jido: ``And where is the other tape?''. In this situation, the
\textsc{Dialogs} module is able to interpret that Jerry is not referring to the
video which they were just talking about, but to the other one:

@@

This experiment involved a more complex decisional layer where the ORO server
was used in conjunction with the HATP symbolic task planner and the SHARY
execution controller (the complete software architecture we deployed for this
experiment is the one pictured on Figure~\ref{fig|archi},
page~\pageref{fig|archi}).

In this scenario (Figure~\ref{fig|cleantable-video}), a human and a robot
cooperate to remove objects from a table. The robot produces symbolic plans for
both itself and the human (Figure~\ref{plan-etat2})
that allow the robot to (verbally) share the task with the human (like ``I take
the green box and I put it in the trashbin, you take the black video tape and
you throw it in the trashbin''). Plans are created based on perceived
visibility and reachability of the objects, and the robot also monitors the
human activities to track the advancement of the whole plan.

@@

Figure~\ref{fig|cleantable-timeline} presents a sample of the whole task and
shows how the different components produce and use symbolic knowledge.

@@

\begin{figure*}[thpb]
  \centering
    \includegraphics[width=1.0\textwidth]{manip_run.pdf} \\
        \caption {Sequence of events of a simplified version of the \emph{Cleaning the table} scenario.}
  \label{fig|cleantable-timeline}
\end{figure*}
